\chapter{Previous Work}

\section{Sensorimotor synchronization}
What follows is a brief overview of SMS. A primer on available tap test software is followed by a detail of critical terminology \ref{SMSTerms}. Last, foundational framework of the current work is established by exploring relevant research within the field \ref{SMSFindings}.

\subsection{Tap Test Software}
Due to the reliability, precision \textit{(ms)}, and discrete nature, the finger tap mechanism holds traditional precedence in SMS research. As such the experiments in this project follow suit. Studies predominantly rely on a MIDI based (drum pad) instrument to register tap events and provide some sort of auditory feedback. A few tap based software suites for experimentation and data acquisition are readily available: a Linux based system written by Finney in 2001 named \textit{FTAP} \cite{finney2001ftap}, and a \textit{Matlab} based toolbox by Elliot in 2009 called \textit{MatTAP} \cite{elliott2009mattap}. FTAP relied on a MIDI source and had a reported mean auditory latency of approximately 14.6ms (SD = 2.8) \cite{schultz2016tap}. Furthermore, superfluous and unregistered taps were common.
Both \textit{FTAP} and \textit{MatTap} were experimented with as viable options for this work but ultimately deemed either outmoded, lacking high baudrate (115200) hardware support for haptic integration over serial, or incompatible with the system architecture in use \textit{(OSX 10.11.6 2.6GHz Intel I7)}.

In a novel high-precision/low-latency approach by Prof. Schultz in 2015 at the University of Montreal \cite{schultz2016tap}, an Arduino force sensitive resistor (FSR) based tap mechanism was constructed. Round-trip latency between tap and auditory feedback was minimized with a mean of 0.6ms (SD = 0.3). The results also demonstrated the reliability of the FSR in recording fewer superfluous taps as well as fewer missed taps.
It was inevitably decided to construct a custom software suite to fit the test needs of the project as discussed in Section \ref{}.

\subsection{Terminology} \label{SMSTerms}
The main method of data collection for SMS tap based tests involve collection of the time delta between the tap and event onset, called the \textit{asynchrony}. The mean of the asynchonies is typically negative (\textit{NMA}), indicative of the participants anticipation of the beat rather than reaction. Positive asynchonies within the shortest reaction time window (150 ms) are arguably an anticipation of the preceeding stimuli. The standard deviation of the asynchrony $(SD_{asy})$ is an index of stability; lower values indicative of a better synchronization.  \cite{repp2013sensorimotor}

Other important metrics include the variability and mean of the inter tap interval (\textit{ITI}) and the inter onset interval (\textit{IOI}), or the time between successive beats - measured in milliseconds. Mismatch between the ITI and IOI implies poor synchronization skill from the participant. 

When a participant is instructed to tap on the beat, this is termed 1:1 synchronization. 4:1 synchronization, for example, is a beat subdivided into four with one tap on the beat. Subdivision tests typically yield lower mean $(SD_{asy})$ values. \cite{repp2013sensorimotor}

\subsection{Findings} \label{SMSFindings}
Professional musicians exhibit a lower ITI variability with percussionists as well as pianists. Surprisingly, both amateurs and non-musicians had no $SD_{asy}$ difference. From a paper presented at the Music Perception and Cognition in 2012:
\begin{quotation}
    Stoklasa, Liebermann, and Fischinger reported that musicians playing their own brass or string instrument in synchrony with a metronome showed a negligible NMA (–2 ms), unlike their tapping (–13 ms). \cite{repp2013sensorimotor}
\end{quotation}

Furthermore, as the duration of the IOI increases, or slower beats per minute, $(SD_{asy})$ increased in a non linear fashion. 

Isochronous vs. nonisochronous ~\cite{polak2016both}


\section{Auditory Advantage}
Decades of research into sensorimotor synchronization presents a clear advantage of the discretely timed auditory stimulus implying that the neural and evolutionary mechanisms underlying beat synchronization are modality-specific.
~\cite{gan2015synchronization} The stability of beat synchronization to discrete visual modalities (a flash of light) has been shown to be less stable that its auditory counterpart.

Concrete examples/figures?

\section{Rhythmic Perception}
Though seemingly a separate realm of study, the field of rhythmic perception is an important contribution to the overall understanding of sensorimotor synchronization. The work involves measurement of the ability to recognize different rhythmic patterns to different stimuli in a listen and respond type of fashion. Researchers from the human computer interaction group at the University of Tampere, Finland, conducted an experiment in 2008 to confirm that the instantaneous auditory modality dominates rhythmic perception. Tactile follows close suit with the visual modality being the least suitable for accurately perceiving rhythmic information as well as the most mentally demanding. Rather than the traditional tap based test, users were given two rhythmic sections and asked to determine whether they were identical or not across modalities as well as combinations of each. ~\cite{jokiniemi2008crossmodal} Even though it yielded less correct results the tactile modality was, from the users point of view, almost as good as the auditory modality. Exploration of pulse length was called upon for further insight.

\section{A Continuous Visual Metronome} \label{visualMet}
In a novel advancement challenging the auditory advantage and perhaps paving the way towards a more meaningful gesture, researchers in the Psychology department at Sun Yat-Sin University in Guangdong found continuous motion of a bouncing ball to be as stable as synchronization to an auditory metronome.
~\cite{gan2015synchronization}


Bouncing ball paper discussion.

\section{The Tactile Modality} \label{tactileModality}
A 2016 study by the Department of Psychology at Ryerson University considered whether the auditory advantage persisted across the tactile modality. The experiment was a tap test of non musicians put through a series of simple and complex rhythmic sequences with a varied area of haptic stimulation. In conditions involving a large area of stimulation and simple rhythmic sequences, tactile synchronization closely matched auditory. They proved that if made salient enough, “the accuracy of synchronization to a tactile metronome can equal synchronization to an auditory metronome, further challenging the idea of an auditory advantage over all other modalities for synchronization to discretely timed rhythmic stimuli.” However, auditory won out for synchronization of complex rhythmic sequences. ~\cite{ammirante2016synchronizing}

\subsection{Multisensory Cues}
\todo[inline]{Revise and reword}
Maintaining synchrony with a periodic event requires that the central nervous system (CNS) compensate for timing variation arising from sensory, decision and motor processing noise. 
Keeping in time with a pacing source (metronome) requires continual corrections based on the timing error (asynchrony) between the metronome and performed actions
The Central Nervous system can alternate between cues depending on the demand of the task or combine info from different senses. In the context of rhythmic cues the brain will weight signals according to the relative reliability in the timing of the events across modalities, ensuring optimal movement production to the underlying event extracted from the signals. 

asynchrony variability for unimodal tactile cues was lower than for the visual metronome (F1,9 = 6.929, P = 0.027) and only slightly higher than that for unimodal auditory cues. \cite{elliott2010multisensory}

\subsection{Haptic Drumkit}
\todo[inline]{Revise and reword}
In 2010, this group at the Open University in the UK first thought of a haptic set which was purposed to enable a drummer to learn multi-limb coordination with the broader goal of polyrhythmic entrainment.

Drew up an important distinction between stimulus response and fostering entrainment.

Notes from test subjects:
	-commented that the haptic guidance was “intimate” and that you didn’t have to work out the division of labor as in audio
	-drumming had a tendency to drown out signal from vibrotactiles - haptic masking
	-Blurring attack of haptics at high tempo
    -for looping patterns it was hard to discern whether the pattern would start
\cite{holland2010feeling}

\subsection{Vibrotactile Metronome} \label{vibrotactileMetronome}
\todo[inline]{Revise and reword}
\cite{ignoto2017development}
The vibrotactile metronome is a thesis project of Patrick Ignoto of the Centre for Interdisciplinary Research in Music Media and Technology (CIRMMT) program at McGill University.

The work done has some fascinating parallels to this project and gave me some very tangible insights towards testing and overall procedure.

Patrick’s overall goal was to propose a device which uses tactile sense to provide similar functionality to a click track as it’s used for a contemporary classical music conductor with the added benefit of not blocking the ear or interfering with the conductors perception.

His guide for design requirements was the director and conductor of the contemp music ensemble, Prof. Bourgogne. He gave him the constraint that the pulses should feel continuous and not discrete, even mentioning a pendulum motion as the descriptive feeling. 

Another constraint was that the pulses peak amplitude line up with the audio track.

The input click track was converted to a vibrotactile click via some Matlab code
	Allowed for simultaneity between audible and vibrotactile pulses

Two ERM’s and one pager buzzer

Transmitter connected to PC running Max initially:
Real-time audio analysis using bonk~ object to find downbeats
Triggered vibrotactile envelope signal and control message trasmitted to device
	Redefined Inter-Onset Interval as half previous IOI (rise time) + half nexts IOI (decay time) for more precision to accommodate varying pulse lengths

Had to move to post-processing in Matlab since couldn’t keep up with buffering
Used findpeaks to find local maxima for each click, determined IOI
To synchronize with the audio click he triggered the haptic pulse midway between two audio clicks


\subsection{Commercial Introspection}
Peterson tuner BodyBeat Sync (\$140) seeks to revolutionize the traditional metronome through its extensive coverage of all three modalities with a wearable pulsing vibration unit which claims to “allow musicians to easily internalize the beat and develop a note value relationship both audibly and physically.” [Peterson Citation]

Ramp up/down as well as proof via quantification of this rhythmic internalization are missing.

The Soundbrenner (\$99) is a vibration based metronome using an instantaneous pulse and claims that in freeing the ears, it has “brought the rhythm closer to the body, making it more comfortable and natural to feel the beat and swing of the music instead of chasing the click.” [Soundbrenner Citation]

Similarly, lack of ramp up/down as well as numerical proof.
